---
title: "Lab 09: Algorithmic Bias"
author: "Kryschelle"
date: "5/04/2026"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(janitor)
library(readr)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r}
# Load the COMPAS data
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )

# Take a look at the data
glimpse(compas)
```

## Exercise Part 1: Exploring the Data

```{r}

nrow(compas)

ncol(compas)

```
The dimensions of the dataset are 7214 by 53. Each row represents a defendant. The variables include demographics and information regarding their charges, offenses, as well as their COMPAS scores and their recidivism rate. 

```{r}

length(unique(compas$name))

```
There are 7158 unique defendants in the dataset. This is not the same as rows because some people may show up multiple times in the dataset due to repeat offenses (or recidivism).

```{r}

ggplot(compas, aes(x = decile_score)) +
  geom_bar()

```

Based on the graph, the distribution is positively skewed, indicating that most people are viewed as low risk. 

```{r}

ggplot(compas, aes(x = race)) +
  geom_bar()

ggplot(compas, aes(x = sex)) +
  geom_bar()

ggplot(compas, aes(x = age)) +
  geom_bar()

```

## Part 2: Risk Scores and Recividism 

```{r}

ggplot(compas, aes(x = factor( decile_score), fill = factor( two_year_recid))) +
  geom_bar(position = "dodge")

```

Higher risk scores don't actually relate to higher levels of recidivism. People who did recidivate two years later did have higher compas scores than people who didn't but people who recidivated don't seem just as likely to have a high risk score as a low risk score. 

```{r}

compas <- compas %>% 
  mutate(classification = case_when(
    decile_score >= 7 & two_year_recid == 1 ~ "TP", 
             decile_score <= 4 & two_year_recid == 0 ~ "TN", 
             decile_score >= 7 & two_year_recid == 0 ~ "FP", 
             decile_score <= 4 & two_year_recid == 1 ~ "FN")
  )

results <- table(compas$classification)
print(results)

correct <- results["TP"] + results["TN"]
cases <- results["TP"] + results["TN"] + results["FP"] + results["FN"]

accuracy <- (correct/cases) * 100
print(accuracy)
```

To see how accurate the test is, I divided the correct cases by the total cases and multipled that by 100 to determine the "accuracy" percentage. About 68% of the cases are correct.

## Part 3: Investigating Disparities 

```{r}

compas_bw <- compas %>%
  filter(race %in% c("African-American", "Caucasian"))

ggplot(compas_bw, aes(x = factor(decile_score), fill = factor(two_year_recid))) +
  geom_bar(position = "dodge") + 
  facet_wrap(~ race) 

compas_bw_b <- compas %>%
  filter(race %in% c("African-American"))

compas_bw_w <- compas %>%
  filter(race %in% c("Caucasian"))

compas_bw_w <- compas_bw_w %>% 
  mutate(classification = case_when(
    decile_score >= 7 & two_year_recid == 1 ~ "TP", 
    decile_score <= 4 & two_year_recid == 0 ~ "TN", 
    decile_score >= 7 & two_year_recid == 0 ~ "FP", 
    decile_score <= 4 & two_year_recid == 1 ~ "FN"
  )) 

res_b <- table(compas_bw_b$classification)
res_w <- table(compas_bw_w$classification)

acc_black <- (res_b["TP"] + res_b["TN"]) / sum(res_b) * 100
acc_white <- (res_w["TP"] + res_w["TN"]) / sum(res_w) * 100


print(acc_black)
print(acc_white)


fp_acc_black <- (res_b["FP"]) / sum(res_b) * 100
fp_acc_white <- (res_w["FP"]) / sum(res_w) * 100

fn_acc_black <- (res_b["FN"]) / sum(res_b) * 100
fn_acc_white <- (res_w["FN"]) / sum(res_w) * 100

print(fp_acc_black)
print(fp_acc_white)

print(fn_acc_black)
print(fn_acc_white)

plot_data <- data.frame(
  Race = c("African-American", "Caucasian", "African-American", "Caucasian"),
  Error_Type = c("False Positive", "False Positive", "False Negative", "False Negative"),
  Percentage = c(fp_acc_black, fp_acc_white, fn_acc_black, fn_acc_white)
)

accuracy_data <- data.frame(
  Race = c("African-American", "Caucasian"),
  Accuracy = c(acc_black, acc_white)
)

print(accuracy_data)

print(plot_data)

ggplot(plot_data, aes(x = Error_Type, y = Percentage, fill = Race)) +
  geom_col(position = "dodge") + 
  labs(
    title = "Comparison of COMPAS Error Rates by Race",
    subtitle = "False Positives vs. False Negatives",
    y = "Percentage (%)",
    x = "Type of Error"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("African-American" = "steelblue", "Caucasian" = "gray"))

ggplot(accuracy_data, aes(x = Race, y = Accuracy, fill = Race)) +
  geom_col() + 
  geom_text(aes(label = round(Accuracy, 1)), visualize = TRUE, vjust = -0.5) +
  labs(
    title = "Overall COMPAS Accuracy by Race",
    subtitle = "Percentage of correct predictions (TP + TN) / Total",
    x = "Race",
    y = "Accuracy Percentage (%)"
  ) +
  scale_fill_manual(values = c("African-American" = "steelblue", "Caucasian" = "gray")) +
  ylim(0, 100) +
  theme_minimal() +
  theme(legend.position = "none") 
```

The overall accuracy is lower for African Americans and the rate of false positives (high risk without recidivation) was higher for African-Americans. Furthermore, way more white individuals were classified as low risk but did recidivate (false negative). Using column plots provides the simplicity of bar and histograms while also modelling relationships. I find bar plots to be some of the simplest to understand when done right and, as such, viewers can easily see the percentages of the accuracy and compare the numbers of false positives and false negatives. 

note: I really enjoy this lab but it is quite long and has taken me a while. It might be useful to condense or remove some sections. 

## Part 4: Understanding the sources of bias

I 

```{r}

ggplot(compas_bw, aes(x = factor(decile_score), y = priors_count, fill = race)) +
  stat_summary(fun = "mean", geom = "bar", position = "dodge") +
  labs(
    title = " Prior Convictions by COMPAS Score",
    x = "Decile Score",
    y = "Priors Count",
    fill = "Race"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("African-American" = "steelblue", "Caucasian" = "gray"))


calibration_data <- compas_bw %>%
  group_by(race, decile_score) %>%
  summarize(
    recidivism_rate = mean(two_year_recid),
    total_people = n()
  ) %>%
  ungroup()

ggplot(calibration_data, aes(x = decile_score, y = recidivism_rate, color = race)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_abline(intercept = 0, slope = 0.1, linetype = "dashed", color = "gray") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "COMPAS Calibration Check by Race",
    subtitle = "Does the same score predict the same recidivism rate for both groups?",
    x = "Decile Score",
    y = "Actual Recidivism Rate",
    color = "Race"
  ) +
  scale_color_manual(values = c("African-American" = "steelblue", "Caucasian" = "gray30")) +
  theme_minimal()
```

Based on the analysis, there is evidence that the algorithm is fair. Obtaining the same score as an African-American predicts similar or the same recidivism rate as for Caucasian's except for around a Decile Score of 3 and 4. However, once the score is high enough, the algorithm can predict the recidivism rate. 

## Part 5: Fairer Algorithms

To make a fairer assessment algorithm, I would mainly address the mischaracterization of Black individuals as high risk. It would be interesting to know what exactly goes into characterizing individuals as high risk versus low risk -- is it the severity of the charge? Is it the felonies? I'm not sure what exactly goes into the algorithm but I would start there. 

When thinking about "fairness" I think people often imagine "equality" instead of "equity." The danger that follows this is an over examination of black behaviors and faster sentencing and labelling of black individuals compared to their white counterparts. We see this happen socially with the adultification of black children. Where a black child gets penalized by the system, their white counterpart may get a warning. As such, black children and people seem to be 'repeat offenders' compared to their white counterparts. Additionally, we have to consider that black people are disporportionately represented in the system. An argument a lot of people present is that more white individuals go to jail. While this is true, people forget porportions. Individuals should consider porportions of the population when redesigning algorithms that can shape people's lives. 
